---
title: "Why Do Employees Leave?"
author: "Minesh Patel"
date: "16 September 2014"
output: html_document
---
### Prerequisite Libraries
```{r Libraries}
library(ggplot2)
```


## Summary
This is an investigation into possible predictor variables to indicate whether 
or not an employee is likely to leave the organisation. The investigation uses 
a sample dataset originally from the Oracle Vision EBS database. This has then
been pre-processed using Oracle BI Applications HR module (version 7.9.6.4). The
methodology used in this investigation can be reproduced on any OBIA HR dataset
and ideally will be modified based on the outcomes. 

The investigation concludes that the most statistically significant predictors
(out of the limited predictors tested) are:

- Years in Service
- Age
- Weekly Hours

The stastitical significance and amplitude of the expected effects are detailed
later, in the Results section.

## Exploratory Analysis
Firstly, the data was loaded. In this case, a CSV extract was used. The SQL for
the extract is found in Appendix 1. This could be extracted directly from the
the Oracle database using ORE.

```{r Data Load}
hrData <- read.csv("HR_Extract_Emp.csv")
```

Additionally, some data tidying is done:

- Converting salary into £k
- Removing employee records with anomalously high salaries
- Converting length of service into years

```{r Data Tidying}
hrData$SALARY <- hrData$SALARY/1000
hrData <- hrData[which(hrData$SALARY < 2000),]
hrData$SERVICE_YEARS <- round(hrData$SERVICE_DAYS/365,2)
hrData$WEEKLY_HOURS <- hrData$HOURS
hrData <- hrData[which(!is.na(hrData$WEEKLY_HOURS)),]
```

Then a logistic regression model was applied using all selected variables:

- Salary
- Years of Service
- Age
- Weekly Hours

```{r Fit of All Variables, warning=FALSE}
model0 <- glm(LEAVE_FLAG ~ SALARY + 
                      SERVICE_YEARS + 
                      AGE + 
                      WEEKLY_HOURS, data = hrData, family = "binomial")
rawCoefs0 <- summary(model0)$coefficients
expCoefs0 <- exp(model0$coefficients)

print(rawCoefs0)
```

The z-test coefficients are compared to a p < 0.1 threshold, deemed suitable for
this study. Only salary has a p value outside of this range with: 
**`r format(rawCoefs0[17], digits=3)`**. Additionally, the (exponential of) the
estimated gradient is: **`r format(expCoefs0[2], digits=3)`**, very close to 1. 
For these reasons it will be excluded from the rest of the study. 

Below is the model applied again, with salary removed.

```{r Best Model Fit, warning=FALSE, message=FALSE}
model1 <- glm(LEAVE_FLAG ~ 
                      SERVICE_YEARS + 
                      AGE + 
                      WEEKLY_HOURS, data = hrData, family = "binomial")
rawCoefs1 <- summary(model1)$coefficients
print(rawCoefs1)
expCoefs1 <- exp(model1$coefficients)
print(expCoefs1)
confInts1 <- exp(confint(model1))
print(confInts1)
```

## Results

We can interpret the coefficients of this model to mean the following:

- For each additional year in service, the employee is 
**`r format((1-expCoefs1[2])*100, digits = 3)` $\pm$
`r format((expCoefs1[2]-confInts1[2])*100, digits = 2)`% less** likely to leave.
- For every year older, the employee is 
**`r format((1-expCoefs1[3])*100, digits = 3)` $\pm$
`r format((expCoefs1[3]-confInts1[3])*100, digits = 2)`% less** likely to leave.
- For every additional weekly hour, the employee is 
**`r format((1-expCoefs1[4])*100, digits = 3)` $\pm$
`r format((expCoefs1[4]-confInts1[4])*100, digits = 2)`% less** likely to leave.

Additionally, the probability of an employee leaving, based on this parameters,
is given by:

```{r Format Coefficients, include=FALSE}
betaF <- 0
betaF[1] <- format(rawCoefs1[1], digits = 2)
betaF[2] <- format(rawCoefs1[2], digits = 2)
betaF[3] <- format(rawCoefs1[3], digits = 2)
betaF[4] <- format(rawCoefs1[4], digits = 2)
```

$$P(x_1, x_2, x_3)=P_L=\frac{e^\gamma}{1 + e^\gamma} \\$$
$$
\begin{align}
\gamma & =\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 \\
& =`r betaF[1]` + `r betaF[2]` x_1 + `r betaF[3]` x_2 + `r betaF[4]` x_3
\end{align}
$$

where $x_1$ is years of service, $x_2$ is age and $x_3$ is the employee's
weekly hours.

### Plot
Below is a plot showing the fit against the observed values.

```{r Plot Fit}
beta0 <- summary(model1)$coefficients[1]
beta1 <- summary(model1)$coefficients[2]
beta2 <- summary(model1)$coefficients[3]
beta3 <- summary(model1)$coefficients[4]

x1 <- hrData$SERVICE_YEARS
x2 <- hrData$AGE
x3 <- hrData$WEEKLY_HOURS

gamma <- beta0 + beta1*x1 + beta2*x2 + beta3*x3

y_fit <- exp(gamma) / (1 + exp(gamma))
y_obs <- hrData$LEAVE_FLAG

plotDF <- data.frame(gamma = gamma, y_obs = y_obs, y_fit = y_fit)
g <-  ggplot() + 
      geom_point(data = plotDF, aes(x = gamma, y = y_obs, col = "1"), 
                 alpha = 0.1) + 
      geom_line(data = plotDF, aes(x = gamma, y = y_fit, col = "2")) + 
      labs(title=expression(bold("Fit against Observed Values")),
           x=expression(beta[0]+x[1]*beta[1]+x[2]*beta[2]+x[3]*beta[3]), 
           y="Probability of Leaving") +
      scale_colour_manual(name = "Plot",
                          values =c('black', 'red'), 
                          labels = c('Observed', 'Fit'))
plot(g)
```

### Prediction Results
The graph above doesn't immediately indicate the effectiveness of the model. The
following compares the effectiveness of the model for employees with $P_L > 0.9$
```{r Prediction Effectiveness}
hrData$P_Leave <- y_fit

p90 <- sum(hrData[which(hrData$P_Leave >= 0.9), "LEAVE_FLAG"])/
      length(hrData[which(hrData$P_Leave >= 0.9), "LEAVE_FLAG"])
p70 <- sum(hrData[which(hrData$P_Leave >= 0.7), "LEAVE_FLAG"])/
      length(hrData[which(hrData$P_Leave >= 0.7), "LEAVE_FLAG"])
```
Out of employees with a predicted leave likelihood of 0.9 or greater, 
**`r format(p90, digits = 3)`**% did in fact leave. Similarly, 
**`r format(p70, digits = 3)`**% left out of the group taken for 0.7 or higher.

## Appendix
### 1 - SQL for data extract
```
select
MAX(Dayd.CALENDAR_DATE) AS CAL_DATE,
MAX(Emp.FULL_NAME) AS FULL_NAME,
Emp.INTEGRATION_ID AS EMP_NUM,
MAX(Wkfc.SALARY_ANNL) AS SALARY, 
MAX(CURRENT_SERVICE_DAYS) AS SERVICE_DAYS, 
MAX(EMP_POW_BAND_DESC) AS PERIOD_OF_WORK_BAND,
MAX(ROUND(AGE_IN_MONTHS/12,0)) AS AGE,
MAX(AGE_BAND_DESC) AS AGE_BAND,
MAX(Wkfc.TERM_EVENT_IND) AS LEAVE_FLAG,
MAX(STD_HOURS_MONTH) AS HOURS,
MAX(Dayd.CALENDAR_DATE - JOB_ENTRY_DT) AS DAYS_IN_JOB,
MAX(Dayd.CALENDAR_DATE - POS_ENTRY_DT) AS DAYS_IN_POS,
MAX(Dayd.CALENDAR_DATE - ORG_ENTRY_DT) AS DAYS_IN_ORG
from
W_WRKFC_EVT_MONTH_F Wkfc,
W_EMPLOYEE_D Emp,
W_DAY_D Dayd,
W_AGE_BAND_D Age_Band,
W_PRD_OF_WRK_BAND_D Period_Wrk
WHERE
Wkfc.EMPLOYEE_WID = Emp.ROW_WID AND
TRUNC(Wkfc.EFFECTIVE_START_DATE) = TRUNC(Dayd.CALENDAR_DATE) AND
Wkfc.AGE_BAND_WID = Age_Band.ROW_WID AND
Wkfc.PRD_OF_WRK_BAND_WID = Period_Wrk.ROW_WID AND
Dayd.CAL_YEAR BETWEEN 2001 AND 2009
GROUP BY 
Dayd.CAL_YEAR,
Emp.INTEGRATION_ID
ORDER BY
Emp.INTEGRATION_ID;
```
